---
title: 'An introduction to the R basics:'
author: "by R Alcala, Y Xing, K Garrett, and A Plex"
date: "12/1/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


Many basic R functions are showed in the present text. 

## R as a calculator

For instance, you can use R as **a simple calculator** for:
1. Addition:
```{r}
1+1
```

2. Multiplication:
```{r}
3*2
```
3. Power:
```{r}
8^2
```
4. Division:
```{r}
7/2
```
Or simply for 
5. Operations order:
```{r}
4+4-2*8
1+(4-2)*8
```

So the result for *4+4-2* is `r 4+4-2`.

## R for objects

R is useful for creating **objects**. The first step is setting values to use again as 'objects. We do this using the assignment operator **<-** (called the *gets arrow*). For example:
```{r}
a <- 2+2
print(a)
a

b <- a+2
b
```

You can also use **<-** for variable assignment:
```{r}
a.vector <- c(4,3,2,1)
x.vector <- mean(other.vector <- c(1,2,3,4))
```
And preferably use **=** to assign parameters:
```{r}
parameter1 = .1
par1 = par2 = par3 = NULL
```

## R for Relational operators

R uses the some symbols to represent relations between two statements *<*, *>*, *==*, *>=*, or *!=* (lesser than, higher than, does equal, greater than or equal to, or does not equal to, respectively). 
1. While comparing values, R will return `r 1 == 1` or `r 1 == 0` responses.
```{r}
TRUE == TRUE
F == F
FALSE != FALSE
```
2. We can also assign the binary system to the `r 1 == 1` or `r 1 == 0` responses:
```{r}
FALSE == 0
TRUE == 1
```
3. Logical numericals:
```{r}
2 == 2 ## '==' means does equal
2 != 2 ## '!=' means does not equal
2 >= 3 ## '>=' means greater than or equal to
```
4. Logical strings:
Assigning the objects with *" "*:
```{r}
apple <- "apple"
pineapple <- "pineapple"
```
So based on the alphabetical order in each word, we might try to compare them.
```{r}
apple == pineapple
apple > pineapple
apple < "abcde"
```

## Creating an object with values from 1 to 10

Well we have to ways to do it in R, see.
```{r}
num10 <- c(1,2,3,4,5,6,7,8,9,10)
num.10 <- 1:10
```
What would be your favorite one?

But now which elements are less than 5 on the above object? Responses will be: `r 1 == 1`s and `r 1 ==0`s.
```{r}
num5 <- num10 < 5
num5
```

Or maybe we just are wishing to subset the objects we created.
```{r}
num10[3:8]
num10[!num5]
```

## Relational Operators & Sub-setting 2D objects

`data()` is a function to call our data.
```{r}
data(iris) # Iris is a famous dataset.
head(iris) # For checking the first rows.
tail(iris) # For checking the last rows.
class(iris) # For checking the classes in the dataset.
str(iris)  # For checking the structure.

iris[1:5, ] # Subset the data: it displays only the first 5 rows [looks like `head()`!]
table(iris$Species)
```

If we wish to select only speals less than 6 cm.
```{r}
low_sl <- iris[iris$Sepal.Length < 6, ]
low_sl
```
Note:  syntax use `data[row, column]`

If we want to create a logical vector for sepal lenght bigger than 5 cm.
```{r}
low_sl$Sepal.Length < 5 # What do you think the result will be?
```

## R for logical operators and ( &), or ( |), not ( !)

For instance, if we select only sepals less than 6 cm **and** with petals less than 6 cm.
```{r}
low_slANDpl <- iris[ c( iris$Sepal.Length  < 6  & iris$Petal.Length <= 6), ]
head(low_slANDpl)
table(low_slANDpl$Species)
```

If we might select only sepals less than 6 cm **or** with petals less than 6 cm.
```{r}
low_slORpl <- iris[ c( iris$Sepal.Length  < 6  | iris$Petal.Length <= 6), ]
head(low_slORpl)
table(low_slORpl$Species)
```

Now, let me ask for a quick question: are they the same species?
```{r message=FALSE, warning=FALSE}
low_slANDpl$Species == low_slORpl$Species
```

If we might selec versicolor species with sepal length below 6 cm.
```{r}
low_slANDsp <- iris[c(iris$Sepal.Length  <  6 & iris$Species == "versicolor"), ]

iris[iris$Sepal.Length != 5,]
```

## Vectors in R

Vectors are a way to set a series of data elements as an object. For example:
```{r}
v1 <- c(1, 2, 3, 4, 5) # numbers
v2 <- c("hello", "world") # characters 
v3 <- c("TRUE", "FALSE", "TRUE") # logical values (also could be "T", "F", "T")
v1
v2
v3
```

## Summary sttistics in R

Lets make a vector with hypothetical ratings of "R expertise2 on a scale of 1-10.
```{r}
WithR <- c(8.5, 6.5, 4, 1, 3, 10, 5, 5, 5, 1, 1, 6, 6)
WithR
```

`summary()`function helps us to look at some summar statistics.
```{r}
summary(WithR)
```

But if you have a doubt about a function you can call `help`.
```{r}
?summary
mean(WithR)
sd(WithR) 
median(WithR)# work inside the vector withinR
vector <- summary(WithR)
vector["Mean"] # accessing the vector for 'Mean'
vector[3] #accessing the vector for position '3'
```

Now, lets plot a basic plot: a histogram or data distribution!
```{r}
hist(WithR, xlab = "Self-Reported R Proficiency")
plot(WithR, xlab = "Self-Reported R Proficiency")
```

## Loading our data in R
First, we set the directory with `setwd("~")`, then we use the read.csv() to search the document.
```{r include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# setwd("~")
# setwd('~/Desktop/')
#Header = TRUE lets R know that we have headings in the first row of our data set.
RustData <- read.csv(file = "StripeRust.csv", header = TRUE)

class(RustData) # let's check the data type 
str(RustData) # let's check the data attributes
head(RustData) ## check the first 6 rows of the dataset
head(RustData, 10) # check the first 10 rows of the dataset
names(RustData) ## check the column headers
```

## Ploting with ggplot

First, lets install the package ggplot and load the library. 
```{r}
library(ggplot2)
```

Now, we can plot this data using ggplot2 (we will learn more about this later!). Tell the graphic funcion to use the data ´RustData´, and add a layer to the plot with `geom_point()`function: severity on the y axis and year on the x axis. Color the points by the year that data was collected. 

```{r}
ggplot(data = RustData) +
  geom_point(mapping = aes(x = DAI, y = Severity, color = Year))

# Switch from 'geom_point' to 'geom_line' we can plot lines and not points.
ggplot(data = RustData) +
  geom_line(mapping = aes(x = DAI, y = Severity, color = Year))+
  theme_classic()

ggplot(data = RustData) +
  geom_point(mapping = aes(x = DAI, y = Severity)) +
  facet_wrap(~Year)
```

## Conditional statements

1. `if` symntax 

 if (test_condition) {
     do_my_condition
 }

```{r}
i <- 2
if (i < 5) {
	message(i, " is less than 5")
}
```

2. `if else` symnyax

 if (test_condition) {
     do_my_condition
   } else {
     do_other_condition}

```{r}
i <- 20
if (i < 5) { # change '<' to '>' and see what happens
	message(i, " is less than 5")
} else {  
	message(i, " is greater than 5")}
```

3. `while` symnyax

 while (test_condition) {
    do_my_condition
    }

```{r}
i <- 1
while (i < 10) {
  print(i)
  } # this is gonna run forever so 'kill the taks' ctrl + c or press 'stop' in your console
```

4. Fullfilling your condition. 

```{r}
while (i < 10) {
  print(i)
  i = i + 1}
```

## Loops in R

1. `for` loop syntax

 for (test_condition) {
    do_my_condition
 }

```{r}
for (i in 1:10){
  print(i)  
  i = i + 1
}
```

2. `while` loop syntax

 for (test_condition) {
    do_my_condition
 }

```{r}
i <- 1
while (i <= 10) {
  print(i)
  i = i + 1
}
```

3. For loops with if else stament

```{r}
for (i in 1:100){ 
  if (i < 50) {
	message(i, " Loops are cool")
  } else {  
	message(i, " Loops are super cool")}
}
```

#End! :) 